Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-04-03 paper.tex
new fig from youdin, incoportated in text

In a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq \epsilon q \hat{z} \Omega^2,
\end{align}
where $\hat{z}$ is a normalized vertical co-ordinate. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

If $\gamma>1$ then the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  For a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \hat{z}^2
  \Omega^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisified in a
thin disk, the disk would be stable.

However, the stabilizing effect of buoyancy can be mitigated for
non-adiabatic disturbances. Consider a fluid element displaced upwards, away from
the disk midplane. In a stably stratified adiabatic disk, it would
sink because of buoyancy. However, if thermal timescales are short, in
that temperature flucations decay rapidly (i.e. isothermal perturbations), then 
the fluid element would attain its surrounding density and 
%(assuming an ideal gas and it reaches pressure 
%equilibrium with its surroundings). 
thus buoyancy is rendered ineffective. Qualitatively, a short thermal
timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail? For a rough
comparison of vertical shear to buoyancy for non-adiabatic
perturbations, we apply Eq. 47 of \cite{urpin03}, which is based on
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk, the destabilizing effect of
vertical shear is rather weak because $|\p_z\Omega|\propto \epsilon
\ll 1$. Then even a small amount of buoyancy can significantly
stabilize the VSI. It follows that $t_c$ must be quite short to permit
instability.

Note that in astrophysical disks both $N_z^2$ and $\p_z\Omega$ varies
with height. Eq. \ref{bcrit_est} then suggest that with 
increasing height, buoyancy would eventually stabilize the VSI (since
$N_z^2\propto z^2$ and $\p_z\Omega\propto z$). 
%This means that 
%disturbances localized away from the disk midplane are easily stabilized by
%buoyancy. 
This argument assumes vertically-localized disturbances. It is,
however, not clear how to apply Eq. \ref{bcrit_est} to vertically
global disturbances, as considered in this work.

@@

Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales in a typical protoplanetary disk model (see 
below). Here, $H$ is the disk scale height and $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}. The criterion Eq. \ref{prelim_bcrit} in fact
assumes $t_c$ is vertically uniform. While this is not generally true
in the inner disk, we find Eq. \ref{prelim_bcrit} still provides a
useful guide. We see that the VSI is likely to  operate in the outer 
disk at tens of AU, but that smaller scale disturbances may operate at
smaller radii.

>>>>

In a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq \epsilon q \Omega^2\left(\frac{z}{H}\right),
\end{align}
where $H$ is the disk scale height. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

If $\gamma>1$ then the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  For a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \Omega^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisified in a
thin disk, the disk would be stable.

However, the stabilizing effect of buoyancy can be mitigated for
non-adiabatic disturbances. Consider a fluid element displaced upwards, away from
the disk midplane. In a stably stratified adiabatic disk, it would
sink because of buoyancy. However, if thermal timescales are short, in
that temperature flucations decay rapidly (i.e. nearly isothermal
perturbations), then the fluid element would attain its surrounding
density and  
%(assuming an ideal gas and it reaches pressure 
%equilibrium with its surroundings). 
thus buoyancy is rendered ineffective. Qualitatively, a short thermal
timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail? For a rough
comparison of vertical shear to buoyancy for non-adiabatic
perturbations, we apply Eq. 47 of \cite{urpin03}, which is based on
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk, the destabilizing effect of
vertical shear is rather weak because $|r\p_z\Omega|\propto \epsilon
\ll 1$. By comparison the buoyancy frequency $N_z/\Omega\sim 1$ at
$z\sim H$.

To illustrate this point, we show in Fig. \ref{eqm_structure} the
equilibrium structure of a typical protoplanetary disk. Here $\rho_0$
is the midplane density and vertical shear rate and buoyancy
frequencies are normalized by the local Keplerian frequency. It is
clear that vertical shear is 
generally weak compared to buoyancy. Then even a small (positive)
vertical entropy gradient can significantly stabilize the VSI. It
follows that $t_c$ must be quite short to permit instability.

\begin{figure*}
  \includegraphics[width=\linewidth]{figures/rhoshearNz}
  \caption{Typical protoplanetary disk structure (the Minimum Mass
    Solar Nebla): density (left), vertical shear rate (middle) and
    buoyancy frequency (right).  
    \label{eqm_structure} 
  }
\end{figure*}

Note that in astrophysical disks both $N_z^2$ and $\p_z\Omega$ varies
with height, as evident in
Fig. \ref{eqm_structure}. Eq. \ref{bcrit_est} then suggest that with
increasing height, buoyancy would eventually stabilize the VSI (since 
$N_z^2\propto z^2$ and $\p_z\Omega\propto z$). 
% This means that 
% disturbances localized away from the disk midplane are easily stabilized by
% buoyancy. 
This argument assumes vertically localized disturbances. It is,
however, not clear how to apply Eq. \ref{bcrit_est} to vertically
global disturbances, as considered in this work.

@@

Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales the protoplanetary disk model shown in
Fig. \ref{eqm_structure}. Here $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}. The criterion Eq. \ref{prelim_bcrit} in fact
assumes $t_c$ is vertically uniform. While this is not generally true
in the inner disk, we find Eq. \ref{prelim_bcrit} still provides a
useful guide. We see that the VSI will operate in the outer 
disk at tens of AU, but that smaller scale disturbances may operate at
smaller radii.
